---
title: '<span style=''font-size: 30px''>**School Protection Brigades power calculations**
  </style> '
author: "Javier Osorio"
date: "v.1.3"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pwr)
```


This power calculations considers two main scenarios.  The first one relies on the universe of public schools of basic education in Mexico City as the pool of units of analysis to conduct the intervention.  The second scenario considers a subset of schools located in hot-spots of violent crime to conduct the intervention.  The analysis below presents the power calculations for both research settings.




# **Full Set of Schools**

### **Assumptions**

The power calculation is based on the following assumptions:

* Each School Protection Brigade (SPB) is considered as a unit of analysis 
* The pool of potential units includes all 2,242 schools in Mexico City
* The perimeter of interest has a radius of 300m around the school
* The treatment and control schools are balanced
* Violent crimes in the perimeter of interest are distributed as:
  * Mean = 48.924
  * Standard Deviation = 46.59
* The menu of violent crimes includes 84 different crime categories.^[The specific violent crimes included in the sutdy are: Sexual Abuse; Sexual Harassment; Breaking and Entering; Threats; Attack on the Roads of Communication (Damage to Roads or Means of Transportation); Attack on the General Ways of Communication; Attacks on the Public Peace; Intentional Damage to Owned Property; Intentional Property Damage to Automobile; Intentional Property Damage to Real Estate Property; Intentional Property Damage to Home Room; Intentional Property Damage to Business; Intentional Damage to Own Property to Ways of Communication; Crimes Against Health (Drug trafficking); Forced Disappearance of Persons; Firearm Shots; Extortion; Feminicide; Homicide by White Gun; Homicide by Firearm; Homicide by Beating; Intentional Homicides (Other); Intimidation; Intentional Injuries; Intentional Injuries by White Weapon; Intentional Injuries by Firearm; Intentional Injuries from Strikes; Intentional Injuries and Vehicle Theft; Drug Possession for Purposes of Sale, Trade and Supply; Simple Drug Possession; Kidnapping; Carrying Prohibited Weapon; Carrying Firearm; Kidnapping and Vehicle Theft; Kidnapping; Kidnapping (To Perform A Sexual Act); Home Robbery Violence; Home Robbery and Vehicle with Violence; Business Robbery with Violence; Theft from Business and Vehicle with Violence; Robbery from Public Office with Violence; Theft of A Passenger / Taxi Driver with Violence; Robbery to Passenger on Board A Metro with Violence; Robbery to Passenger on Board A Metrobus with Violence; Robbery to Passenger on Board A Collective Bus with Violence; Robbery to Passenger on Board A Taxi with Violence; Robbery to Passenger on Board of Public Transportation with Violence; Robbery to Passenger on A Foreign Bus with Violence; Robbery to Passenger in Ecobus with Violence; Theft to Passenger in RTP with Violence; Robbery to Passenger in A Trolebus with Violence; Robbery to Dealer with Violence; Theft to Dealer and Vehicle with Violence; Theft to Bank Branch (Bank Assault) with Violence; Theft to Bank Branch (Supermarket) with Violence; Theft to Cell Phone with Violence; Robbery to Pedestrian in Hotel with Violence; Robbery to Pedestrian in Business with Violence; Robbery to Pedestrian in Parks and Markets with Violence; Robbery to A Restaurant with Violence; Robbery to Passenger in Passenger Terminal with Violence; Robbery to Bystanding on Public Road with Violence; Robbery to Pedestrian Leaving the Bank with Violence; Robbery to Passenger Leaving the Atm with Violence; Robbery to Pedestrian and Vehicle with Violence; Theft from Carrier and Heavy Vehicle with Violence; Theft of Machinery with Violence; Motorcycle Theft with Violence; Theft of a Transportation Service Vehicle with Violence; Theft of a Particular Service Vehicle with Violence; Theft of a Public Service Vehicle with Violence; Theft Inside a Company (Payroll) with Violence; Child Abduction; Extorsion Attempt; Homicide Attempt; Attempted Robbery; Vehicle Theft Attempt; Attempted Violation; Torture; Trafficking in Persons; Violation; Equipped Rape; Gang Rape; and Domestic Violence.]  
* The SSC will have to review and approve the selection of schools before advancing into implementation 




### **Power calculations** 

The power analysis considers the following scenarios:

\

#### **Case 1.a Weak Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 4.6 less violent crimes
* Corresponds to a 10% decline

```{r}
# Sample size for a given power
p.out1 <-power.t.test(power=0.8,
                      delta=4.6,
                      sd=46.59,
                      type="two.sample")
p.out1
```

To detect an effect of -4.6 crimes with a probability of 80%, the study needs `r round(p.out1$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out1$n*2, digits=0)` schools.

Unfortunately, the costs associated with covering such a large number of schools does not make this option feasible. 


\

#### **Case 2.a Moderate Weak Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 7.3 less violent crimes
* Corresponds to a 15% decline

```{r}
# Sample size for a given power
p.out2 <-power.t.test(power=0.8,
                      delta=7.3,
                      sd=46.59,
                      type="two.sample")
p.out2
```

To detect an effect of -7.3 crimes with a probability of 80%, the study needs `r round(p.out2$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out2$n*2, digits=0)` schools.

Unfortunately, the costs associated with covering such a large number of schools does not make this option feasible. 




\

#### **Case 3.a Moderate Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 9.3 less violent crimes
* Corresponds to a 20% decline

```{r}
# Sample size for a given power
p.out3 <-power.t.test(power=0.8,
                      delta=9.3,
                      sd=46.59,
                      type="two.sample")
p.out3
```

To detect an effect of -9.3 crimes with a probability of 80%, the study needs `r round(p.out3$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out3$n*2, digits=0)` schools.



\

#### **Case 4.a Moderate Strong Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 12.2 less violent crimes
* Corresponds to a 25% decline

```{r}
# Sample size for a given power
p.out4 <-power.t.test(power=0.8,
                      delta=12.2,
                      sd=46.59,
                      type="two.sample")
p.out4
```


To detect an effect of -12.2 crimes with a probability of 80%, the study needs `r round(p.out4$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out4$n*2, digits=0)` schools.




\

#### **Case 5.a Strong Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 13.9 less violent crimes
* Corresponds to a 30% decline

```{r}
# Sample size for a given power
p.out5 <-power.t.test(power=0.8,
                      delta=13.9,
                      sd=46.59,
                      type="two.sample")
p.out5
```


To detect an effect of -13.9 crimes with a probability of 80%, the study needs `r round(p.out5$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out5$n*2, digits=0)` schools.



\

\newpage

### **Power against sample size for Full Set of Schools**

The following graph presents the power calculation and the optima number of observations for scenarios 3, 4, and 5. This excludes scenarios 1 and 2, which are not feasible due to their large number of schools.

```{r fig.height = 4, fig.width = 5, fig.align = "center", warning=FALSE}
# Get power calculatons
samplesizes <- seq(from=0,to=800,by=10)
#pwr.1 <-power.t.test(n=samplesizes,delta=4.6  ,sd=46.59,type="two.sample")$power
#pwr.2 <-power.t.test(n=samplesizes,delta=7.3  ,sd=46.59,type="two.sample")$power
pwr.3 <-power.t.test(n=samplesizes,delta=9.3  ,sd=46.59,type="two.sample")$power
pwr.4 <-power.t.test(n=samplesizes,delta=12.2 ,sd=46.59,type="two.sample")$power
pwr.5 <-power.t.test(n=samplesizes,delta=13.9 ,sd=46.59,type="two.sample")$power

# Generate plot
plot(samplesizes, pwr.3, type="b", col="skyblue1", lwd=2, axes=FALSE, ylab="", xlab="" )
par(new=TRUE)
plot(samplesizes, pwr.4, type="b", col="dodgerblue", lwd=2, axes=FALSE, ylab="", xlab="" )
par(new=TRUE)
plot(samplesizes, pwr.5,type="b", col="mediumblue", lwd=2, axes=FALSE,
     xlim=c(0,800), xlab="Sample size", ylab="Expected power", ylim=c(0,1))
axis(1,at=c(0,200,400,600,800))
axis(2,at=c(0,0.2,0.4,0.6,0.8,1),labels=paste(c(0,20,40,60,80,100),"%"))
abline(h=0.8, col="red", lty=2)
abline(v=p.out3$n, col="skyblue1", lty=2)
abline(v=p.out4$n, col="dodgerblue", lty=2)
abline(v=p.out5$n, col="mediumblue", lty=2)
legend(500, 0.3, legend=c("delta=13.9", "delta=12.2", "delta=9.3"),
       col=c("mediumblue", "dodgerblue", "skyblue1"), lwd=2:2, lty=1:1, cex=0.8, box.lty=0)
```





### **Summary from Full Set of Schools**

Based on the power calculations, the following scenarios indicate the number of schools necessary to detect with 80% probability an effect within a radius of 300m.


| Effect   | % effect | # Crimes | Schools per group | Total Schools | Feasible |
|----------|:----------:|:----------:|:-------------:|:---------:|:---------:|
| Weak                | -10%     | -4.6     | `r round(p.out1$n, digits=0)` | `r round(p.out1$n*2, digits=0)`    | No |
| Moderate Weak       | -15%     | -7.3     | `r round(p.out2$n, digits=0)` | `r round(p.out2$n*2, digits=0)`    | No |
| Moderate            | -20%     | -9.3     | `r round(p.out3$n, digits=0)` | `r round(p.out3$n*2, digits=0)`    | Yes |
| Moderate Strong     | -25%     | -12.2    | `r round(p.out4$n, digits=0)` | `r round(p.out4$n*2, digits=0)`    | Yes |
| Strong              | -30%     | -13.9    | `r round(p.out5$n, digits=0)` | `r round(p.out5$n*2, digits=0)`    | Yes |




### **Selected Design for Full Set of Schools**

Based in the previous power calculation, the project will pursue the following research design:


```{r}
# Effect for a given power and sample size
design.1 <-power.t.test(power=0.8,
                      n=300,
                      sd=46.59,
                      type="two.sample")
```

* **Number of schools in each group = `r round(design.1$n, digits=0)`**
* **Total number of schools = `r round(design.1$n*2, digits=0)`**
* **Power = `r round((design.1$power)*100, digits=2)`%**
* **Minimum Detectable Effect (MDE) = `r round(design.1$delta, digits=0)` less violent crimes**
* **MDE proportion = -`r round((design.1$delta/46.59)*100, digits=2)`%**

This research design represents the best possible balance between maximizing the probability of detecting a relatively moderate effect while keeping the project within the realm of cost feasibility, time of implementation, and operational capacity in the field.





\ 


\newpage


# **Subset of Schools in Hotspot Perimeters**

In order to increase the probability of detecting an effect, the research design will also consider focusing the RCT implementation on violent crime hot-spots.

Of course, the SSC will have to review this possibility and approve the selection of hotspot schools before advancing into implementation.



### **Assumptions**

The power calculation is based on the following assumptions:

* All previous assumptions hold
* Now we focus on schools in hotspot areas
* Hotspot is defined as an perimeter of 300 m concentrating more than the average number of violent crimes 
* This subset includes a total of 859 schools
* Violent crimes in hotspot perimeters are distributed as:
  * Mean = 86.43
  * Standard Deviation = 55.73









### **Power calculations for hot-spots** 

The power analysis considers the following scenarios:

\

#### **Case 1.b Weak Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 8.64 less violent crimes
* Corresponds to a 10% decline

```{r}
# Sample size for a given power
p.out1.h <-power.t.test(power=0.8,
                      delta=(86.43*.1),
                      sd=55.73,
                      type="two.sample")
p.out1.h
```

To detect an effect of -8.64 crimes with a probability of 80%, the study needs `r round(p.out1.h$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out1.h$n*2, digits=0)` schools.

Unfortunately, the costs associated with covering such a large number of schools does not make this option feasible. 


\

#### **Case 2.b Moderate Weak Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 12.96 less violent crimes
* Corresponds to a 15% decline

```{r}
# Sample size for a given power
p.out2.h <-power.t.test(power=0.8,
                      delta=(86.43*.15),
                      sd=55.73,
                      type="two.sample")
p.out2.h
```

To detect an effect of -12.96 crimes with a probability of 80%, the study needs `r round(p.out2.h$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out2.h$n*2, digits=0)` schools.





\

#### **Case 3.b Moderate Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 17.29 less violent crimes
* Corresponds to a 20% decline

```{r}
# Sample size for a given power
p.out3.h <-power.t.test(power=0.8,
                      delta=(86.43*.2),
                      sd=55.73,
                      type="two.sample")
p.out3.h
```

To detect an effect of -17.29 crimes with a probability of 80%, the study needs `r round(p.out3.h$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out3.h$n*2, digits=0)` schools.



\

#### **Case 4.b Moderate Strong Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 21.61 less violent crimes
* Corresponds to a 25% decline

```{r}
# Sample size for a given power
p.out4.h <-power.t.test(power=0.8,
                      delta=(86.43*.25),
                      sd=55.73,
                      type="two.sample")
p.out4.h
```


To detect an effect of -21.61 crimes with a probability of 80%, the study needs `r round(p.out4.h$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out4.h$n*2, digits=0)` schools.




\

#### **Case 5.b Strong Effect**

Sample size for a given power:

* Power = 80%
* Intervention effect = 25.93 less violent crimes
* Corresponds to a 30% decline

```{r}
# Sample size for a given power
p.out5.h <-power.t.test(power=0.8,
                      delta=(86.43*.3),
                      sd=55.73,
                      type="two.sample")
p.out5.h
```


To detect an effect of -25.93 crimes with a probability of 80%, the study needs `r round(p.out5.h$n, digits=0)` schools in each group.

Which corresponds to a total number of `r round(p.out5.h$n*2, digits=0)` schools.



\

\newpage

### **Power against sample size for Hotspot Schools**

The following graph presents the power calculation and the optima number of observations for scenarios 3, 4, and 5. This excludes scenarios 1 and 2, which are not feasible due to their large number of schools.

```{r fig.height = 4, fig.width = 5, fig.align = "center", warning=FALSE}
# Get power calculatons
samplesizes <- seq(from=0,to=800,by=10)
#pwr.1 <-power.t.test(n=samplesizes,delta=(86.43*.1)  ,sd=55.73,type="two.sample")$power
pwr.2.h <-power.t.test(n=samplesizes,delta=(86.43*.15) ,sd=55.73,type="two.sample")$power
pwr.3.h <-power.t.test(n=samplesizes,delta=(86.43*.2)   ,sd=55.73,type="two.sample")$power
pwr.4.h <-power.t.test(n=samplesizes,delta=(86.43*.25)  ,sd=55.73,type="two.sample")$power
pwr.5.h <-power.t.test(n=samplesizes,delta=(86.43*.3)   ,sd=55.73,type="two.sample")$power

# Generate plot
plot(samplesizes, pwr.2.h, type="b", col="cadetblue1", lwd=2, axes=FALSE, ylab="", xlab="" )
par(new=TRUE)
plot(samplesizes, pwr.3.h, type="b", col="skyblue1", lwd=2, axes=FALSE, ylab="", xlab="" )
par(new=TRUE)
plot(samplesizes, pwr.4.h, type="b", col="dodgerblue", lwd=2, axes=FALSE, ylab="", xlab="" )
par(new=TRUE)
plot(samplesizes, pwr.5.h,type="b", col="mediumblue", lwd=2, axes=FALSE,
     xlim=c(0,800), xlab="Sample size", ylab="Expected power", ylim=c(0,1))
axis(1,at=c(0,200,400,600,800))
axis(2,at=c(0,0.2,0.4,0.6,0.8,1),labels=paste(c(0,20,40,60,80,100),"%"))
abline(h=0.8, col="red", lty=2)
abline(v=p.out2.h$n, col="cadetblue1", lty=2)
abline(v=p.out3.h$n, col="skyblue1", lty=2)
abline(v=p.out4.h$n, col="dodgerblue", lty=2)
abline(v=p.out5.h$n, col="mediumblue", lty=2)
legend(500, 0.4, legend=c("delta=25.93", 
                          "delta=21.61", 
                          "delta=17.29",
                          "delta=12.96"),
       col=c("mediumblue", 
             "dodgerblue", 
             "skyblue1",
             "cadetblue1"), lwd=2:2, lty=1:1, cex=0.8, box.lty=0)
```





### **Summary from Hotspot Schools**

Based on the power calculations, the following scenarios indicate the number of schools necessary to detect with 80% probability an effect within a radius of 300m.


| Effect   | % effect | # Crimes | Schools per group | Total Schools | Feasible |
|----------|:----------:|:----------:|:-------------:|:---------:|:---------:|
| Weak                | -10%     | -8.64     | `r round(p.out1.h$n, digits=0)` | `r round(p.out1.h$n*2, digits=0)`    | No |
| Moderate Weak       | -15%     | -12.96     | `r round(p.out2.h$n, digits=0)` | `r round(p.out2.h$n*2, digits=0)`    | Yes |
| Moderate            | -20%     | -17.29     | `r round(p.out3.h$n, digits=0)` | `r round(p.out3.h$n*2, digits=0)`    | Yes |
| Moderate Strong     | -25%     | -21.61    | `r round(p.out4.h$n, digits=0)` | `r round(p.out4.h$n*2, digits=0)`    | Yes |
| Strong              | -30%     | -25.93    | `r round(p.out5.h$n, digits=0)` | `r round(p.out5.h$n*2, digits=0)`    | Yes |



### **Selected Design for Hotspot Schools**

Based in the previous power calculation, the project will considering the research design based on case 2:


```{r}
# Effect for a given power and sample size
design.2 <-power.t.test(power=0.8,
                      n=300,
                      sd=55.73,
                      type="two.sample")
```

* **Number of schools in each group = `r round(design.2$n, digits=0)`**
* **Total number of schools = `r round(design.2$n*2, digits=0)`**
* **Power = `r round((design.2$power)*100, digits=2)`%**
* **Minimum Detectable Effect (MDE) = `r round(design.2$delta, digits=0)` less violent crimes**
* **MDE proportion = -`r round((design.2$delta/86.43)*100, digits=2)`%**

This research design represents the best possible balance between maximizing the probability of detecting a moderately weak effect while keeping the project within the realm of cost feasibility, time of implementation, and operational capacity in the field.





\ 


\newpage


# **Alternative Approaches**

To increase the probability of detecting an effect, the research team could consider the following approaches:

1. **Consider a different specification of hot-spots**: 

    * Define hot-spots of violent crime with a different threshold or located in other priority neighborhoods.
    * Currently, several government programs in Mexico City provide focalized attention to a set of areas called the "333 neighborhoods." 
    * In coordination with the SSC, tThe project could consider focusing on those areas.
    * The set of hot-spots, would constitute the population from which we could draw the sample of schools for the intervention. 
    * This assumes that schools in crime hot-spots have a higher mean and a smaller standard deviation than the ones considered here.
    * This alternative will be explored in further detail in consultation with the Mexico City police in order to maximize the methodological rigor of the project with their strategic priorities.
    
2. **Consider 911 calls**: 

    * The analysis could consider 911 calls as indicator to assess the impact of the intervention. 
    * This assumes that the number of 911 calls related to violent crimes is higher than the number of crimes reported in official statistics.


